Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-12-13 manu-draft.qmd
proofing manu draft

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services[@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable, if not more, than other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007], whereas natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create supportive conditions for growth. A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014] . There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014]and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in areal coverage through sustained and long-term reductions in nutrients. Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016]. In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016]. Costly alternative management actions, in addition to nutrient reductions, are needed in these situations. Management actions are much more effective in estuaries that are stressed but not past critical thresholds that define an alternative state.

@@

In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017]. Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres (16,857 ha) [@greening2006; @greening2014]. Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006]. Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth. Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers. Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay, although at a relatively coarse scale [@nickerson2023]. Changing salinity has not been well-described, nor have potential links of changing salinity and temperature with recent seagrass change. Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions. This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay Estuary over the last fifty years. Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that conditions in Tampa Bay are trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges. This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys. Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable. Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery. The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.

@@

Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico. The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay. Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer months. The watershed is heavily developed and includes over 3 million people with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023]. The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a]. Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management efforts have historically focused on restoring and maintaining seagrass cover [@sherwood2017]. Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barrens) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20]. Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 188,429 acres [76,254 ha, @beck2023].

@@

Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s [e.g., @SWFWMD23]. These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December. The TBEP has used these maps to track progress towards achieving seagrass restoration goals as total cover in Tampa Bay. No species information is provided. A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 61 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information, including abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the transect data allow for inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Additional sources of seagrass data are described in the next section.

@@

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b). These data provide the basis for regulatory assessments and compliance reporting for the nutrient TMDL in Tampa Bay. Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements for physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Time of sampling can vary, although most samples are collected from mid-morning to early afternoon. All surface and bottom salinity and temperature measurements for each of the 45 monitoring stations were evaluated herein. Trends were assessed for both surface and bottom samples, as described below, whereas only the bottom measurements were used for comparison to seagrass trends. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

The second dataset used to evaluate water quality trends was available from the Florida Fish and Wildlife Conservation Commission (FWC). The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple survey gear types that target different habitats. A stratified sample design is used to target multiple habitats where unique sites are sampled each month. We used data from the 21.3 meter center-bag nearshore seine that targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c). These data are collected by pulling the seine adjacent to the shore to sample approximately 140 m$^2$ of bay bottom [@schrandt2021]. In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the beginning and end of the seine haul, and typically at the surface and bottom. Only measurements at the beginning of the seine haul and from the bottom were used. Seagrass data are also provided for each site, with information on species and cover. Total percent cover for all species at a site was used for comparison with temperature and salinity measurements. Sites exclusively with macroalgae were not included in the analysis. All FIM data were provided from FWC staff by request.

The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM). Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting. Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (also available at <https://wateratlas.usf.edu/>). We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments (Figure [-@fig-map]d). Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone. Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained. Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.

@@

Formal trend tests were used to assess station-level changes in water temperature and salinity in the EPC data. These analyses also provided a detailed spatial assessment of trends because the EPC data is the only dataset of the three where the same sites have been sampled over time. Seasonal Kendall trend tests were used to evaluate the monotonic change for temperature and salinity from 1975 to present at each water quality station [@hirsch1982; @millard13]. The change per year was also evaluated for each parameter based on the slope estimates returned by each test. Kendall tests were also used to evaluate changes over time for each month to determine when the trends were most pronounced seasonally, e.g., all January estimates across years, all February estimates, etc. The percentage of stations in each bay segment with significant increasing temperature or decreasing salinity trends were evaluated for each month. All tests evaluated both surface and bottom measurements to assess potential differences by water depth.

@@

The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity, with particular attention on differences between bay segments and the time periods before and after 2016 (pre/post seagrass recovery). The conceptual model for evaluating these changes describes the niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972]. In the simplest form, this can conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges. Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space [@fig-concept]. Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum range, or when both temperature and salinity conditions are outside of the optimum range. Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right @fig-concept). Although the optimal niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have generally not been limiting for growth in recent years.

@@

Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to both describe the potential for stress and how it may be related to changes in seagrass. Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay. First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold. This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may are may not be outside of the optimum range for seagrasses. These metrics were quantified from the monthly long-term observations in the EPC data. To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity and a single predictor for decimal year [@wood17]. The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends. Model fit for each station was sufficient to calculate daily predictions for assessing the stressor metrics (see supplement, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models).

@@

For comparison to seagrass, the annual metrics were also referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year as above. Transect surveys are typically done in the fall at the end of the growing season. The average date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly. For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates. This was repeated for all years from 1998 to 2022 when transect data were available. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment. The stressor metrics were compared to frequency occurrence (all species) each year by bay segment. Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016). Two models were evaluated, one with the temperature and salinity metrics and another with the both metric given that the latter was a combined metric of the former. Interactions were included between temperature, salinity, and time period or the both metric and time period depending on the model. Bay segment was included as a main effect without interactions. The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time.

Additional models were constructed for the FIM and PDEM datasets to evaluate seagrass change relative to temperature and salinity. These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs. GLMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables. Interactions with time period and a main effect for bay segment were also included as above. GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrass. Input data for models for the FIM and PDEM datasets were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.

@@

Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation [@fig-meteowqraw]. Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$C. Similarly, total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`). Total precipitation in 1975 was `r prpstr`, whereas current total precipitation in 2022 was `r prpend`, showing an overall increase in the period of record of `r prpdif` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record. The SPI showed notable anomalies in precipitation, with pronounced rainy periods in 1980, the late 1990s, 2005, and the last ten years.

>>>>

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable, if not more, than other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007], whereas natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create supportive conditions for growth. A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014]. There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in nutrients. Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016]. In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016]. Costly alternative management actions, in addition to nutrient reductions, are needed in these situations. Management actions are much more effective in estuaries that are stressed but not past critical thresholds that define an alternative state.

@@

In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017]. Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014]. Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006]. Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth [@tbep0696; @Beck20a]. Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers. Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay, although at a relatively coarse scale [@nickerson2023]. Changing salinity has not been well-described, nor have potential links of changing salinity and temperature with recent seagrass change. Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions. This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay estuary over the last fifty years. Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that conditions in Tampa Bay are trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges. This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring describing cover of individual species, and synoptic seagrass data collected with the water quality surveys. Specific focus was on the periods prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable. Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery. The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems at the tropical-temperate boundary, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.

@@

Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico. The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee Rivers contributing a majority of freshwater inflow to the bay. Tampa Bay straddles the tropical and temperate boundary of central Florida characterized by warm, humid conditions and a distinct rainy season during the summer. The watershed is heavily developed and includes over 3 million people with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023]. The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a]. Important subtidal habitats include seagrasses, hard bottom, tidal flats, and oyster reefs, where a majority of management efforts have historically focused on restoring and maintaining seagrass cover [@sherwood2017]. Additional native habitats include intertidal wetlands (mangroves, salt marshes, salt barrens) and pine forests, oak hammocks, and freshwater wetlands present in upland habitats [@Robison20]. Losses of native uplands and potentially restorable habitats to development in the watershed from 1990 to 2020 have been estimated at 188,429 acres [76,254 ha, @beck2023].

@@

Two primary sources of data have been used to track seagrass change over time in Tampa Bay. The Southwest Florida Water Management District has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>). These maps are created by photointerpretation of aerial images obtained at the end of the growing season, typically during November-December. No species information is provided. A more detailed, but spatially-specific, data source is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information on abundance, cover, frequency occurrence, and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the transect data allow inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Additional sources of seagrass data are described in the next section.

@@

Several datasets with distinct sample designs are available to assess long-term changes in water temperature and salinity in Tampa Bay. These datasets were evaluated individually to assess trends and relationships with seagrass change to provide a weight-of-evidence approach for potential causal relationships driving the recent decline. First, the Environmental Protection Commission (EPC) of Hillsborough County has collected discrete water quality measurements monthly at 45 stations in the major bay segments since the late 1970s (Figure [-@fig-map]b). These data are used for regulatory assessments for the nutrient TMDL in Tampa Bay. Water quality samples are collected at each station from surface water grabs (e.g., nutrients) or *in situ* measurements of physical parameters (e.g., salinity, temperature) collected at the surface, mid-depth, and bottom. Time of sampling can vary, although most samples are collected from mid-morning to early afternoon. All surface and bottom salinity and temperature measurements for each of the 45 monitoring stations were evaluated herein. Trends were assessed for both surface and bottom samples, as described below, whereas only the bottom measurements were used for comparison to seagrass trends. The data were obtained using the tbeptools R package that imports the data directly from a stable web address provided by the EPC [@beck2021].

The second dataset used to evaluate water quality trends was available from the Florida Fish and Wildlife Conservation Commission (FWC). The Fisheries Independent Monitoring (FIM) program administered by FWC provides monthly surveys of the entire nekton community in Tampa Bay, including species richness and abundance, using multiple survey gear types that target different habitats. A stratified sample design is used to target multiple habitats at different sites sampled each month. We used data from the 21.3 meter center-bag nearshore seine that targets shallow habitats where seagrasses are predominantly found in Tampa Bay and includes the longest consistent sampling protocol (1996 to present, Figure [-@fig-map]c). These data are collected by pulling the seine adjacent to the shore to sample approximately 140 m$^2$ of bay bottom [@schrandt2021]. In addition to collecting fish and selected invertebrates, *in situ* physical measurements for water temperature and salinity are collected at the beginning and end of the seine haul, and typically at the surface and bottom. Only measurements at the beginning of the seine haul and from the bottom were used. Seagrass data are also provided for each site, with information on species and cover. Total percent cover for all species at a site was used for comparison with temperature and salinity measurements. Sites exclusively with macroalgae were not included in the analysis. All FIM data were provided from FWC staff by request.

The third and final dataset evaluated was from the Pinellas County Department of Environmental Management (PDEM). Surface waters in Pinellas County have been monitored since the 1990s, although a consistent stratified random sampling designed has been used in Tampa Bay since 2003 primarily to support robust statistical assessments for NPDES reporting. Data were obtained by request to PDEM staff for the western portion of Old Tampa Bay where sampling occurs from 2003 to present (Figure [-@fig-map]d, also available at <https://wateratlas.usf.edu/>). We focused primarily on OTB for the analysis of the PDEM data given the length of record, consistency of sampling, and relative loss of seagrass compared to the other bay segments. Four distinct spatial zones in OTB are used to stratify the random selection of sample points that typically include 4 sample points per month in each zone. Water quality samples at each site are similar to those collected by EPC, where only bottom measurements for salinity and temperature were retained for analysis. Seagrass presence/absence is also recorded at each site and all sites were defined as "seagrass" if only seagrass species were identified (any with macroalgae were excluded) and "no seagrass" if bare sediment was observed.

@@

Formal trend tests were used to assess station-level changes in water temperature and salinity in the EPC data. These analyses also provided a detailed spatial assessment of trends because the EPC data is the only dataset of the three where the same sites have been sampled over time. Seasonal Kendall trend tests were used to evaluate the monotonic change for temperature and salinity from 1975 to present at each water quality station [@hirsch1982; @millard13]. The change per year was also evaluated for each parameter based on the slope estimates returned by each test. Kendall tests were also used to evaluate changes over time for each month across years to determine when the trends were most pronounced seasonally, e.g., all January estimates across years, all February estimates, etc. The percentage of stations in each bay segment with significantly increasing temperature or decreasing salinity trends were evaluated for each month. All tests evaluated both surface and bottom measurements to assess potential differences by water depth.

@@

The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity, with particular attention on differences between bay segments and the time periods before and after 2016 (pre/post seagrass recovery). The conceptual model for evaluating these changes describes the niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972]. In the simplest form, this can conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges. Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space ([@fig-concept]). Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum range, or when both temperature and salinity conditions are outside of the optimum range. Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right, @fig-concept). Although the optimal niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have generally not been limiting for growth in recent years.

@@

Because of the uncertainty in defining thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to describe the potential for stress and how it may be related to changes in seagrass. Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay. First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold. This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may are may not be outside of the optimum range for seagrasses. These metrics were quantified from the monthly long-term observations in the EPC data. To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity with a single predictor for decimal year [@wood17]. The smoothing spline for decimal year had sufficient knots to capture the seasonal signal within each year and the long-term inter-annual trends. Model fit for each station was sufficient to calculate daily predictions to assess potential stressor metrics (see supplement, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models).

@@

For comparison to seagrass, the annual metrics were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends. Transect surveys are typically done in the fall at the end of the growing season. The average sampling date for each year across the subset of transects in each bay segment was estimated and the stressor metrics were referenced accordingly. For example, if the average transect date for a bay segment was September 15th in 2009 and Oct 1st in 2010, the 2010 stressor counts were based on all days between the two average dates. This was repeated for all years from 1998 to 2022 when transect data were available. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment. The stressor metrics were compared to frequency occurrence (all species) each year by bay segment. Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016). Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric given that the latter was a combined metric of the former. Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model. Bay segment was included as a main effect without interactions. The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time.

Additional models were constructed for the FIM and PDEM datasets to evaluate seagrass change relative to temperature and salinity. These models used direct measurements of salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs. GLMs were used for the FIM data to evaluate seagrass percent cover using continuous temperature and salinity as independent variables. Interactions with time period and a main effect for bay segment were also included as above. GLMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated (no bay segment interaction) and the seagrass response variable was presence/absence (i.e., a binomial distribution was used) to model the probability of observing seagrass. Model input data were further subset to include only months from July to November to describe seagrasses during the growing season and to reduce potential seasonal effects.

@@

Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation ([@fig-meteowqraw]). Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr` $^\circ$C, whereas current mean annual air temperature in 2022 was `r airtempend` $^\circ$C, showing an overall increase in the period of record of `r airtempdif` $^\circ$C. Similarly, total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`). Total precipitation in 1975 was `r prpstr` mm, whereas current total precipitation in 2022 was `r prpend` mm, showing an overall increase in the period of record of `r prpdif` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record. The SPI showed notable anomalies in precipitation, with pronounced rainy periods in the early 1980s, late 1990s, 2005, and the last ten years.

Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement). Note that for tables [-@tbl-temptrndtab] and [-@tbl-saltrndtab], comparable time periods were evaluated between the datasets when possible given the different sample sizes, and therefore power, to detect trends. The strongest trends were observed for the EPC dataset, which had the longest record from 1975 to 2022. The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths and minimal stratification in Tampa Bay (@fig-meteowqraw). Trends in water temperature were similar across bay segments with significant increases varying from 0.03 to 0.04 $^\circ$C per year, with a total increase of 1.4 (OTB) to 2.0 (HB) $^\circ$C from 1975 to present (@tbl-temptrndtab). Salinity trends were similar between bay segments, although overall salinity was of course higher for bay segments closer to the Gulf of Mexico. Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, with a total decrease of 2.7 (OTB) and 2.0 (LTB) ppt from 1975 to present. As expected, intra-annual variability of water temperature was much higher than salinity following the larger seasonal variation in temperature (@fig-meteowqraw). This variation may also be related to differences among the monitoring stations in each segment because the averages combine monthly and spatial variation. Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in OTB and HB (@tbl-temptrndtab). No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present nor from 2004 to present (see figures in supplement). However, the EPC dataset showed significant increasing temperature across all time periods and bay segments. Salinity trends from the EPC dataset were not significant from 1996 to present, but were significantly decreasing from 2004 to present in all bay segments (@tbl-saltrndtab).

The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall). All stations had significantly increasing temperature and decreasing salinity from 1975 to present in both the top and bottom of the water column, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a). Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed between bay segments. Temperature trends were more often observed in the summer, early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends. Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments. Again, differences between top or bottom salinity trends were not observed.
